\documentclass[12pt]{article}
\usepackage{lecture}
\usepackage{graphics}
\usepackage{epstopdf}
\usepackage{html}
\usepackage{url}

\newcommand{\copyrightYears}{2010-2015}

\title{Association mapping: the background from two-locus genetics}

\begin{document}

\maketitle

\thispagestyle{first}

\section*{Introduction}

One approach to understanding more about the genetics of quantitative
traits takes advantage of the increasing number of genetic markers
available as a result of recent advances in molecular
genetics. Suppose you have two inbred lines that differ in a trait
that interests you, say body weight or leaf width. Call one of them
the ``high'' line and the other the ``low''
line.\footnote{Corresponding to whether the body weight or leaf width
  is large or small.} Further suppose that you have a whole bunch of
molecular markers that differ between the two lines, and designate the
genotype in the ``high'' line $A_1A_1$ and the genotype in the low
line $A_2A_2$.\footnote{Since these are inbred lines, I can assume
  that they are homozygous at the marker loci I've chosen.} One last
supposition: Suppose that at loci influencing the phenotype you're
studying the genotype in the ``high'' line is $Q_1Q_1$ and the
genotype in the ``low'' line is $Q_2Q_2$. Each of these loci is what
we call a {\it quantitative trait locus\/} or QTL.\index{quantitative
  trait locus}\index{QTL} Now do the following experiment:

\begin{itemize}

\item Cross the ``high'' line and the ``low'' line to construct an
  $F_1$. 

\item Intercross individuals in the $F_1$ generation to form an
  $F_2$.\footnote{Note: You could also backcross to either or both of
    the parental inbred lines. Producing an $F_2$, however, allows you
    to estimate both the additive and dominance effects associated
    with each QTL.}

\item ``Walk'' through the genome\footnote{I forgot to mention another
    supposition. I am supposing that you either have already
    constructed a genetic map using your markers, or that you will
    construct a genetic map using segregation in the $F_2$ before you
    start looking for QTL loci.} calculating a likelihood score for a
  QTL at a particular map position, using what we know about the
  mathematics of recombination rates and Mendelian genetics. In
  calculating the likelihood score we maximize the likelihood of the
  data assuming that there is a QTL at this position and estimating
  the corresponding additive and dominance effects of the allele. We
  then identify QTLs as those loci where there are ``significant''
  peaks in the map of likelihood scores.\footnote{See
    \url{http://darwin.eeb.uconn.edu/eeb348/lecture-notes/qtl-intro.pdf}
    for more details.}

\end{itemize}
The result is a genetic map showing where QTLs are in the genome and
indicating the magnitude of their additive and dominance effects.

QTL mapping is wonderful{\dash}provided that you're working with an
organism where it's possible to design a breeding program and where
the information derived from that breeding program is relevant to
variation in natural populations. Think about it. If we do a QTL
analysis based on segregation in an $F_2$ population derived from two
inbred lines, all we really know is which loci are associated with
phenotypic differences {\it between those two lines}. Typically what
we really want to know, if we're evolutionary biologists, is which
loci are associated with phenotypic differences {\it between
  individuals in the population we're studying}. That's where
association mapping comes in. We look for statistical associations
between phenotypes and genotypes across a whole population. We expect
there to be such associations, if we have a dense enough map, because
some of our marker loci will be closely linked to loci responsible for
phenotypic variation.

\section*{A digression into two-locus population
  genetics\footnote{{\bf Note}: We'll go over only a small part of
  this section in lecture. I'm providing all the details here so you
  can find them in the future if you ever need them.}}

It's pretty obvious that if two loci are closely linked, alleles at
those loci are likely to be closely linked, but let's take a closer
look at exactly what that means. 

One of the most important properties of a two-locus system is that it
is no longer sufficient to talk about allele frequencies alone, even
in a population that satisfies all of the assumptions necessary for
genotypes to be in Hardy-Weinberg proportions at each locus. To see
why consider this. With two loci and two alleles there are four
possible gametes:\footnote{Think of drawing the Punnett square for a
  dihybrid cross, if you want.}\index{two-locus genetics!gamet frequenies}

\begin{center}
\begin{tabular}{lcccc}
Gamete    & $A_1B_1$ & $A_1B_2$ & $A_2B_1$ & $A_2B_2$ \\
Frequency & $x_{11}$ & $x_{12}$ & $x_{21}$ & $x_{22}$
\end{tabular}
\end{center}

If alleles are arranged randomly into gametes then,
\begin{eqnarray*}
x_{11} &=& p_1p_2 \\
x_{12} &=& p_1q_2 \\
x_{21} &=& q_1p_2 \\
x_{22} &=& q_1q_2 \quad ,
\end{eqnarray*}
where $p_1 = \hbox{freq}(A_1)$ and $p_2 = \hbox{freq}(A_2)$. But
alleles need not be arranged randomly into gametes. They may covary so
that when a gamete contains $A_1$ it is more likely to contain $B_1$
than a randomly chosen gamete, or they may covary so that a gamete
containing $A_1$ is less likely to contain $B_1$ than a randomly
chosen gamete. This covariance could be the result of the two loci
being in close physical association, but it doesn't have to
be. Whenever the alleles covary within gametes
\begin{eqnarray*}
x_{11} &=& p_1p_2 + D \\
x_{12} &=& p_1q_2 - D \\
x_{21} &=& q_1p_2 - D \\
x_{22} &=& q_1q_2 + D \quad ,
\end{eqnarray*}
where $D = x_{11}x_{22} - x_{12}x_{22}$ is known as the {\it gametic
  disequilibrium}.\footnote{You will sometimes see $D$ referred to as
  the linkage disequilibrium, but that's misleading. Alleles at
  different loci may be non-randomly associated even when they are not
  linked.} When $D \ne 0$ the alleles within gametes covary, and $D$
measures {\it statistical\/} association between them. It does not
(directly) measure the {\it physical\/} association. Similarly, $D =
0$ does not imply that the loci are unlinked, only that the alleles at
the two loci are arranged into gametes independently of one
another.\index{gametic disequilibrium}\index{two-locus genetics!gametic disequilibrium}\index{linkage disequilibrium}

\subsection*{A little diversion}

It probably isn't obvious why we can get away with only one $D$ for
all of the gamete frequencies. The short answer is:
\begin{quote}There are four gametes. That means we need three
  parameters to describe the four frequencies. $p_1$ and $p_2$ are
  two. $D$ is the third.
\end{quote}
Another way is to do a little algebra to verify that the definition is
self-consistent.
\begin{eqnarray*}
D &=& x_{11}x_{22} - x_{12}x_{21} \\
  &=& (p_1p_2 + D)(q_1q_2 + D) - (p_1q_2 - D)(q_1p_2 - D) \\
  &=& \left(p_1q_1p_2q_2 + D(p_1p_2 + q_1q_2) + D^2\right) \\
  && \quad - \left(p_1q_1p_2q_2 - D(p_1q_2 + q_1p_2) + D^2\right) \\
  &=& D(p_1p_2 + q_1q_2 + p_1q_2 + q_1p_2) \\
  &=& D\left(p_1(p_2 + q_2) + q_1(q_2 + p_2)\right) \\
  &=& D(p_1 + q_1) \\
  &=& D \quad. 
\end{eqnarray*}

\section*{Transmission genetics with two loci}

I'm going to construct a reduced version of a mating table to see how
gamete frequencies change from one generation to the next. There are
ten different two-locus genotypes (if we distinguish coupling,
$A_1B_1/A_2B_2$, from repulsion, $A_1B_2/A_2B_1$, heterozygotes as we
must for these purposes). So a full mating table would have 100
rows. If we assume all the conditions necessary for genotypes to be in
Hardy-Weinberg proportions apply, however, we can get away with just
calculating the frequency with which any one genotype will produce a
particular gamete.\footnote{We're assuming random union of {\it
    gametes\/} rather than random mating of {\it
    genotypes}.}\index{two-locus genetics!transmission}

\begin{center}
\begin{tabular}{lccccc}
\hline\hline
         &           & \multicolumn{4}{c}{Gametes} \\
Genotype & Frequency & $A_1B_1$ & $A_1B_2$ & $A_2B_1$ & $A_2B_2$ \\
\hline
$A_1B_1/A_1B_1$ & $x_{11}^2$ & 1 & 0 & 0 & 0 \\
$A_1B_1/A_1B_2$ & $2x_{11}x_{12}$ & $\half$ & $\half$ & 0 & 0 \\
$A_1B_1/A_2B_1$ & $2x_{11}x_{21}$ & $\half$ & 0 $\half$ & 0 \\
$A_1B_1/A_2B_2$ & $2x_{11}x_{22}$ & \oneminus & \rhalf & \rhalf & \oneminus \\
$A_1B_2/A_1B_2$ & $x_{12}^2$ & 0 & 1 & 0 & 0 \\
$A_1B_2/A_2B_1$ & $2x_{12}x_{21}$ & \rhalf & \oneminus & \oneminus & \rhalf \\
$A_1B_2/A_2B_2$ & $2x_{12}x_{22}$ & 0 & $\half$ & 0 & $\half$ \\
$A_2B_1/A_2B_1$ & $x_{21}^2$ & 0 & 0 & 1 & 0 \\
$A_2B_1/A_2B_2$ & $2x_{21}x_{22}$ & 0 & 0 & $\half$ & $\half$ \\
$A_2B_2/A_2B_2$ & $x_{22}^2$ & 0 & 0 & 0 & 1 \\
\hline
\end{tabular}
\end{center}

\subsection*{Where do $\frac{1-r}{2}$ and $\frac{r}{2}$ come from?}

Consider the coupling double heterozygote, $A_1B_1/A_2B_2$. When
recombination doesn't happen, $A_1B_1$ and $A_2B_2$ occur in equal
frequency ($1/2$), and $A_1B_2$ and $A_2B_1$ don't occur at all. When
recombination happens, the four possible gametes occur in equal
frequency ($1/4$). So the recombination frequency,\footnote{The
  frequency of recombinant gametes in double heterozygotes.} $r$, is
half the crossover frequency,\footnote{The frequency of cytological
  crossover during meiosis.} $c$, i.e., $r = c/2$. Now the results of
crossing over can be expressed in this table:\index{two-locus genetics!recombination}\index{recombination frequency}

\begin{center}
\begin{tabular}{c|cccc}
\hline\hline
Frequency & $A_1B_1$ & $A_1B_2$ & $A_2B_1$ & $A_2B_2$ \\
\hline
$1-c$     & $\half$  & 0        & 0        & $\half$ \\
$c$       & $\quarter$ & $\quarter$ & $\quarter$ & $\quarter$ \\
\hline
Total     & $\frac{2-c}{4}$ & $\frac{c}{4}$ & $\frac{c}{4}$
          & $\frac{2-c}{4}$ \\
          & $\frac{1-r}{2}$ & $\frac{r}{2}$ & $\frac{r}{2}$
          & $\frac{1-r}{2}$ \\          
\hline
\end{tabular}
\end{center}

\subsection*{Changes in gamete frequency}

We can use the mating table table as we did earlier to calculate the
frequency of each gamete in the next generation. Specifically,
\begin{eqnarray*}
x_{11}' &=& x_{11}^2 + x_{11}x_{12} + x_{11}x_{21} + (1-r)x_{11}x_{22}
            + rx_{12}x_{21} \\
        &=& x_{11}(x_{11} + x_{12} + x_{21} + x_{22})
            - r(x_{11}x_{22} - x_{12}x_{21}) \\
        &=& x_{11} - rD \\
x_{12}' &=& x_{12} + rD \\
x_{21}' &=& x_{21} + rD \\
x_{22}' &=& x_{22} - rD \quad .
\end{eqnarray*}

\subsection*{No changes in allele frequency}

We can also calculate the frequencies of $A_1$ and $B_1$ after this
whole process:
\begin{eqnarray*}
p_1' &=& x_{11}' + x_{12}' \\
     &=& x_{11} - rD + x_{12} + rD \\
     &=& x_{11} + x_{12} \\
     &=& p_1 \\
p_2' &=& p_2 \quad .
\end{eqnarray*}
Since each locus is subject to all of the conditions necessary for
Hardy-Weinberg to apply at a single locus, allele frequencies don't
change at either locus. Furthermore, genotype frequencies at each
locus will be in Hardy-Weinberg proportions. But the two-locus gamete
frequencies change from one generation to the next.\index{two-locus genetics!Hardy-Weinberg}

\subsection*{Changes in $D$}

You can probably figure out that $D$ will eventually become zero, and
you can probably even guess that how quickly it becomes zero depends
on how frequent recombination is. But I'd be astonished if you could
guess exactly how rapidly $D$ decays as a function of $r$.  It takes a
little more algebra, but we can say precisely how rapid the decay will
be.\index{two-locus genetics!decay of disequilibrium}
\begin{eqnarray*}
D' &=& x_{11}'x_{22}' - x_{12}'x_{21}' \\
   &=& (x_{11} - rD)(x_{22} - rD) - (x_{12} + rD)(x_{21} + rD) \\
   &=& x_{11}x_{22} - rD(x_{11} + x_{12}) + r^2D^2
       - (x_{12}x_{21} + rD(x_{12} + x_{21}) + r^2D^2) \\
   &=& x_{11}x_{22} - x_{12}x_{21} - rD(x_{11} + x_{12} + x_{21} + x_{22}) \\
   &=& D - rD \\
   &=& D(1-r)
\end{eqnarray*}
Notice that even if loci are unlinked, meaning that $r = 1/2$, $D$
does not reach 0 immediately. That state is reached only
asymptotically. The two-locus analogue of Hardy-Weinberg is that
gamete frequencies will {\it eventually\/} be equal to the product of
their constituent allele frequencies.

\subsection*{$D$ in a finite population}

In the absence of mutation, $D$ will eventually decay to 0, although
the course of that decay isn't as regular as what I've just
shown~\cite{Hill-Robertson-1968}. If we allow recurrent mutation at
both loci, however, where\index{two-locus genetics!drift}\index{gametic disequilibrium!drift}
\[
\begin{array}{ccccccc}
    &\mu_1            &     &      &     &\mu_2 \\
A_1 &\rightleftharpoons& A_2 &\qquad& B_1 &\rightleftharpoons& B_2
\quad , \\
    &\nu_1            &     &      &     &\nu_2 
\end{array}
\]
then it can be shown~\cite{Ohta-Kimura-1969} that the expected value
of $D^2/p_1(1-p_1)p_2(1-p_2)$ is 
{\scriptsize
\begin{eqnarray*}
\frac{\mbox{E}(D^2)}{\mbox{E}(p_1(1-p_1)p_2(1-p_2))}
&=& \frac{1}{3 + 4N_e(r+\mu_1+\nu_1+\mu_2+\nu_2)
                           - \frac{2}{(2.5 + N_e(r+\mu_1+\nu_1+\mu_2+\nu_2)
                              + N_e(\mu_1+\nu_1+\mu_2+\nu_2))}} \\
&\approx& \frac{1}{3 + 4N_er} \quad .
\end{eqnarray*}
}
\noindent Bottom line: In a finite population, we don't expect $D$ to
go to 0, and the magnitude of $D^2$ is inversely related to amount of
recombination between the two loci. The less recombination there is
between two loci, i.e., the smaller $r$ is, the larger the value of
$D^2$ we expect.

This has all been a long way\footnote{OK. You can say it. A {\bf\it
    very\/} long way.} of showing that our initial intuition is
correct. If we can detect a statistical association between a marker
locus and a phenotypic trait, it suggests that the marker locus and a
locus influence expression of the trait are physically linked. So how
do we detect such an association and why do I say that it {\it
  suggests} the loci are physically linked?

\section*{Population structure with two loci}

You can probably guess where this is going. With one locus I showed
you that there's a deficiency of heterozygotes in a combined sample
even if there's random mating within all populations of which the
sample is composed. The two-locus analog is that you can have gametic
disequilibrium in your combined sample even if the gametic
disequilibrium is zero in all of your constituent
populations. Table~\ref{table:d-structure} provides a simple numerical
example involving just two populations in which the combined sample
has equal proportions from each population.\item{Wahlund effect!two loci}

\begin{table}
\begin{center}
\begin{tabular}{c|cccc|cc|c}
\hline\hline
           & \multicolumn{4}{c|}{Gamete frequencies} 
           & \multicolumn{2}{c|}{Allele frequencies} \\
Population & $A_1B_1$ & $A_1B_2$ & $A_2B_1$ & $A_2B_2$ 
           & $p_{i1}$ & $p_{i2}$ & $D$ \\
\hline
1          & 0.24     & 0.36     & 0.16    & 0.24
           & 0.60     & 0.40     & 0.00 \\
2          & 0.14     & 0.56     & 0.06    & 0.24
           & 0.70     & 0.20     & 0.00 \\
Combined   & 0.19     & 0.46     & 0.11    & 0.24
           & 0.65     & 0.30     & -0.005 \\
\hline
\end{tabular}
\end{center}
\caption{Gametic disequilibrium in a combined population
  sample.}\label{table:d-structure}
\end{table}

\subsection*{The gory details}

You knew that I wouldn't be satisfied with a numerical example, didn't
you? You knew there had to be some algebra coming, right? Well, here
it is. Let
\begin{eqnarray*}
D_i &=& x_{11,i} - p_{1i}p_{2i} \\
D_t &=& \bar x_{11} - \bar p_1\bar p_2 \quad ,
\end{eqnarray*}
where $\bar x_{11} = \frac{1}{K} \sum_{k=1}^K x_{11,k}$, $\bar p_1 =
\frac{1}{K} \sum_{k=1}^K p_{1k}$, and $\bar p_2 = \frac{1}{K}
\sum_{k=1}^K p_{2k}$. Given these definitions, we can now caclculate
$D_t$. 
\begin{eqnarray*}
D_t &=& \bar x_{11} - \bar p_1\bar p_2 \\
    &=& \frac{1}{K} \sum_{k=1}^K x_{11,k} - \bar p_1\bar p_2 \\
    &=& \frac{1}{K} \sum_{k=1}^K (p_{1k}p_{2k} + D_k) - \bar p_1\bar p_2 \\
    &=& \frac{1}{K} \sum_{k=1}^K (p_{1k}p_{2k} - \bar p_1\bar p_2) + \bar D \\
    &=& \mbox{Cov}(p_1, p_2) + \bar D \quad ,
\end{eqnarray*}
where $\mbox{Cov}(p_1, p_2)$ is the covariance in allele frequencies
across populations and $\bar D$ is the mean within-population gametic
disequilibrium. Suppose $D_i = 0$ for all subpopulations. Then $\bar D
= 0$, too (obviously). But that means that
\begin{eqnarray*}
D_t &=& \hbox{Cov}(p_1, p_2) \quad .
\end{eqnarray*}
So if allele frequencies covary across populations, i.e.,
$\mbox{Cov}(p_1, p_2) \ne 0$, then there will be non-random
association of alleles into gametes in the sample, i.e., $D_t \ne 0$,
even if there is random association alleles into gametes within each
population.\footnote{Well, duh! Covariation of allele frequencies
  across populations means that alleles are non-randomly associated
  across populations. What other result could you possibly expect?}

Returning to the example in Table~\ref{table:d-structure}
\begin{eqnarray*}
\mbox{Cov}(p_1, p_2) &=& 0.5(0.6-0.65)(0.4-0.3) + 0.5(0.7-0.65)(0.2-0.3) \\
                     &=& -0.005 \\
\bar x_{11}          &=& (0.65)(0.30) - 0.005 \\
                     &=& 0.19 \\
\bar x_{12}          &=& (0.65)(0.7) + 0.005 \\
                     &=& 0.46 \\
\bar x_{21}          &=& (0.35)(0.30) + 0.005 \\
                     &=& 0.11 \\
\bar x_{22}          &=& (0.35)(0.70) - 0.005 \\
                     &=& 0.24 \quad .
\end{eqnarray*}

\section*{Association mapping}

So what does any of this have to do with QTL mapping? Imagine that we
have a well-mixed population segregating both for a lot of molecular
markers spread throughout the genome and for loci influencing a trait
we're interested in, like body weight or leaf width. Let's call our
measurement of that trait $z_i$ in the $i$th individual. Let $x_{ij}$
be the genotype of individual $i$ at the $j$th locus.\footnote{To keep
  things simple I'm assuming that we're dealing with biallelic loci,
  e.g., SNPs, and we can then order the genotypes as 0, 1, 2 depending
  on how many copies of the most frequent allele they carry.} Then to
do association mapping, we simply fit the following regression model:
\[
y_i = x_{ij}\beta_j + \epsilon_{ij} \quad ,
\]
where $\epsilon_{ij}$ is the residual error in our regression estimate
and $\beta_j$ is our estimate of the effect of substituting one allele
for another at locus $j$, i.e., the additive effect of an allele at
locus $j$. If $\beta_j$ is significantly different from 0, we have
evidence that there is a locus linked to this marker that influences
the phenotype we're interested in.

Notice that I claimed we have evidence that the locus is
linked. That's a bit of sleight of hand. What we have evidence for
directly is that the locus is {\it associated}. As we've just seen,
though, that association could reflect population structure rather
than physical linkage. So in practice the regression model we fit is a
bit more complicated than the one I showed. The simplest case is when
individuals fall into obvious groups, e.g., samples from different
populations. Then $y_i^{(k)}$ is the trait value for individual
$i$. The superscript $(k)$ indicates that this individual belongs to
group $k$.
\[
y_i^{(k)} = x_{ij}\beta_j + \phi^{(k)} + \epsilon_{ij} \quad .
\]
The difference between this model and the one above is that we include
a random effect of group, $\phi^{(k)}$, to account for the fact that
individuals may have similar phenotypes not because of similarity in
genotypes at loci close to those we've scored but because of their
similarity at other loci that differ among groups.

\bibliography{popgen}
\bibliographystyle{plain}

\ccLicense

\end{document}



