\documentclass[12pt]{article}
\usepackage{lecture}
\usepackage{html}
\usepackage{graphicx}
\usepackage{epstopdf}

\newcommand{\copyrightYears}{2001-2023}

\title{Genetic transmission in populations}

\begin{document}

\maketitle

\thispagestyle{first}

\section*{Introduction}

Transmission genetics is exemplified by Mendel's rules, which describe
how genetic information is transmitted from parents to
offspring. Consider a monohybrid cross:

\begin{center}
\begin{tabular}{ccc}
\multicolumn{3}{c}{$A_1A_2$ $\times$ $A_1A_2$} \\
 & $\downarrow$ & \\
$\frac{1}{4}A_1A_1$ & $\frac{1}{2}A_1A_2$ & $\frac{1}{4}A_2A_2$ \\
\end{tabular}
\end{center}

\noindent Population genetics describes how genetic information is
transmitted from a {\it population\/} of parents to a population of
offspring.  Consider, for example, the following data from the {\it
  Est\/}-3 locus of {\it Zoarces viviparus}:\footnote{Only one
  offspring from each mother is included in these data
  (from~\cite{Christiansen-1980}).}\index{Zoarces
  viviparus@\textit{Zoarces viviparus}}

\begin{center}
\begin{tabular}{lrrr}
                  & \multicolumn{3}{c}{Genotype of offspring} \\
Maternal genotype & $A_1A_1$ & $A_1A_2$ & $A_2A_2$ \\
\hline
$A_1A_1$          &      305 &      516 & \\
$A_1A_2$          &      459 &     1360 & 877 \\
$A_2A_2$          &          &      877 & 1541 \\
\end{tabular}
\end{center}

\noindent This table describes, empirically, the relationship between
the genotypes of mothers and the genotypes of their offspring. Using
these data we can also make some inferences about the genotypes of the
fathers in this population, even though we didn't collect
them.\footnote{As you'll see, we can't actually say much about {\it
    fathers}, but we can make inferences about the genes they
  contributed to the offspring.}

\begin{enumerate}

\item 305 out of 821 male gametes that fertilized eggs from $A_1A_1$
mothers carried the $A_1$ allele (37\%).

\item 877 out of 2418 male gametes that fertilized eggs from $A_2A_2$
mothers carried the $A_1$ allele (36\%).

\end{enumerate}

\begin{description}

\item[Question] How many of the 2,696 male gametes that fertilized
eggs from $A_1A_2$ mothers carried the $A_1$ allele?

\item[Recall] We don't know the paternal genotypes or we wouldn't be
asking this question.

\begin{itemize}

\item There is no way to tell which of the 1360 $A_1A_2$ offspring
  received $A_1$ from their mother and which from their father.

\item Regardless of what the genotype of the father is, half of the
  offspring of a heterozygous mother will be
  heterozygous.\footnote{Assuming we're looking at data from a locus
    that has only two alleles. If there were four alleles at a locus,
    for example, {\it all\/} of the offspring could be
    heterozygous. If you don't see that immediately, think about an
    $A_1A_2$ mother mating with an $A_3A_4$ father and write out the
    Punnett square. We are also assuming that meiosis is fair, i.e.,
    that there's no segregation distortion, and that there's no gamete
    competition and no gamete-gamete assortative mating. If you don't
    know what any of that means don't worry about it. We'll get to
    (most of) it over the next few weeks.}

\item Heterozygous offspring of heterozygous mothers contain no
information about the frequency of $A_1$ among fathers, so we don't
bother to include them in our calculations.

\end{itemize}

\item[Rephrase] How many of the 1336 homozygous progeny of
heterozygous mothers received an $A_1$ allele from their father?

\item[Answer] 459 out of 1336 (34\%)

\item[New question] How many of the offspring where the paternal
contribution can be identified received an $A_1$ allele from their
father?

\item[Answer] (305 + 459 + 877) out of (305 + 459 + 877 + 516 + 877 +
1541) or 1641 out of 4575 (36\%)

\end{description}

\section*{An algebraic formulation of the problem}

The above calculations tell us what's happening for this particular
data set, but those of you who know me know that there has to be a
little math coming so that we can describe the situation more
generally.\footnote{Those of you who don't know me will soon come to
  expect that there is (almost always) more math coming. But don't get
  scared. I don't expect you to reproduce the match. I show it to you
  so that you understand where the tools we will use come from.}
Here's the notation:

\begin{center}
\begin{tabular}{ccr}
\hline\hline
Genotype & Number & Sex \\
\hline
$A_1A_1$ & $F_{11}$ & female \\
$A_1A_2$ & $F_{12}$ & female \\
$A_2A_2$ & $F_{22}$ & female \\
$A_1A_1$ & $M_{11}$ & male \\
$A_1A_2$ & $M_{12}$ & male \\
$A_2A_2$ & $M_{22}$ & male \\
\hline
\end{tabular}
\end{center}

\noindent Using that notation, 

$$\begin{array}{cc}
p_f = \frac{2F_{11}+F_{12}}{2F_{11}+2F_{12}+2F_{22}} &
q_f = \frac{2F_{22}+F_{12}}{2F_{11}+2F_{12}+2F_{22}} \\
 & \\
p_m = \frac{2M_{11}+M_{12}}{2M_{11}+2M_{12}+2M_{22}} &
q_m = \frac{2M_{22}+M_{12}}{2M_{11}+2M_{12}+2M_{22}} \quad ,
\end{array}$$
where $p_f$ is the frequency of $A_1$ in mothers and $p_m$ is the
frequency of $A_1$ in fathers.\footnote{$q_f = 1 - p_f$ and $q_m = 1 -
  p_m$ as usual.}

Since every individual in the population must have one father and one
mother, the frequency of $A_1$ among offspring is the same in both
sexes, namely
\[
p = \frac{1}{2}(p_f + p_m) \quad ,
\]
assuming that all matings have the same average fecundity and that the
locus we're studying is autosomal.\footnote{And that there are enough
  offspring produced that we can ignore genetic drift. Have you
  noticed that I have a fondness for footnotes? You'll see a lot more
  before the semester is through, and you'll soon discover that most
  of my weak attempts at humor are buried in them.}

{\bf Question}: Why do those assumptions matter?

{\bf Answer}: If $p_f = p_m$, then the allele frequency among
offspring is equal to the allele frequency in their parents, i.e., the
allele frequency doesn't change from one generation to the next. This
might be considered the First Law of Population Genetics: {\it If no
  forces act to change allele frequencies between zygote formation and
  breeding, allele frequencies will not change.}

\subsection*{Zero force laws}\index{zero force laws}

This is an example of what philosophers call a {\bf zero force
law}. Zero force laws play a very important role in scientific
theories, because we can't begin to understand what a force does until
we understand what would happen in the absence of any forces. Consider
Newton's famous dictum: 
\begin{quotation}
\noindent An object in motion tends to remain in motion in a straight
line. An object at rest tends to remain at rest.
\end{quotation}
or (as you may remember from introductory physics)\footnote{Don't
  worry if you're not good at physics. I'm probably worse. What I'm
  telling you now is almost the only thing about physics I can
  remember. I graduated from college with only one semester of college
  physics, and it didn't even require calculus as a pre-requisite. My
  brother is an electrical engineer, and he is appalled by my
  inability to remember Ohm's Law.}
\[
F = ma \quad.
\]
\noindent If we observe an object accelerating, we can immediately
infer that a force is acting on it. Not only that, we can also infer
something about the magnitude of the force.  {\bf However}, if an
object is not accelerating we cannot conclude that no forces are
acting. It might be that opposing forces act on the object in such a
way that the resultant is no {\it net\/} force. Acceleration is a {\it
  sufficient\/} condition to infer that force is operating on an
object, but it is not {\it necessary}.

What we might call the ``First Law of Population Genetics'' is
analogous to Newton's First Law of Motion:\index{First law of
  population genetics}
\begin{quotation}
\noindent If all genotypes at a particular locus have the same average
fecundity and the same average chance of being included in the breeding
population, allele frequencies in the population will remain constant
from one generation to the next.
\end{quotation}
For the rest of the semester we'll be learning about the processes
that cause allele frequencies to change and learning how to infer the
properties of those processes from the changes that they induce. But
you must always remember that while we can infer that some
evolutionary process is happening if allele frequencies change from
one generation to the next, we {\it cannot\/} infer the absence of an
evolutionary process from a lack of allele frequency
change.\footnote{If you've been paying very close attention, you will
  have noticed that I changed from talking about ``forces'' to talking
about ``evolutionary processes.'' There's a reason for that. One
important evolutionary process, genetic drift, isn't a
force. Merriam-Webster defines force as ``strength or energy exerted
or brought to bear.'' While natural selection can be thought of as a
force, since it ``pushes'' a population in a particular direction,
genetic drift can't be thought of as a force, since it describes the
random change of a population that happens because it is small and
when the change doesn't have a directional tendency.}

\bibliography{popgen}
\bibliographystyle{plain}

\ccLicense

\end{document}


