---
title: "GWAS in gypsy moths"
output: html_notebook
---

# GWAS with `Stan`

Friedline et al.^[Friedline, C.J., T.M. Faske, B.M. Lind, E.M. Hobson, D. Parry, R.J. Dyer, D.M. Johnson, L.M. Thompson, K.L. Grayson, and A.J. Eckert.  2019.  Evolutionary genomics of gypsy moth populations sampled along a latitudinal gradient. _Molecular Ecology_ 28:2206-2223 doi: [https://doi.org/10.1111/mec.15069](https://doi.org/10.1111/mec.15069)] investigated the role of natural selection in the rapid spread of gypsy moth (_Lymantria dispar_) across Eastern North America. Their analysis focused on analysis of polymorphisms at 91,468 SNPs identified through double digest restriction-site associated DNA sequencing (ddRADseq). As part of their work, they also collected data on pupal mass (Mass), pupal development time (PD), and total development time (TDT). This allows us to perform a genome-wide association analysis that assesses the relationship, if any, between individual SNP loci and each of the phenotypes. You’ll find the data set on the course web site:

- [gypsymoth.csv](http://darwin.eeb.uconn.edu/eeb348-resources/gypsymoth.csv) A CSV file containing information about each individual in the sample (population of origin, sample label, phenotype, genotype at each SNP locus).

The data set is a subset of the data included in the paper. Specifically, I filtered the data to include only (a) loci that were scored in more than 100 individuals, (b) individuals that had more than 4000 of the remaing SNP loci scored, and (c) loci that were scored in all of the remaining individuals. The resulting data set includes 141 individuals scored at 218 loci. Using these data, answer the following questions:

1. What loci (if any) are associated with each of the three traits (pupal mass, pupal development time, and total development time)?

IMPORTANT NOTE: The loci are ordered from the strongest estimated effect (mean) to the weakest, but loci showing a weaker estimated effect may have stronger evidence for them, i.e., the credible interval may not overlap zero. Because of the small sample size, I recommend considering both the 80% and the 95% credible intervals whenn answering these questions. 

In the example below you'll see that X4293 has the largest estimated effect, 0.01434 and that that the 95% credible interval on this effect does not overlap zero. This suggests that genotypic differences at X4293 have an effect on pupal mass. Notice that the 80% credible intervals for X2840, X4680, X2841, and X5330 do not overlap zero, providing suggestive evidence that differences at these loci may also affect pupal mass. Estimated effects at remaining loci are smaller and the 80% credible intervals indicating that we cannot detect any effect differences at these loci have on pupal mass.^[It's important to be very careful in your conclusion here. We do ***not*** have evidence that differences at these loci have no effect. With a larger sample of individuals we might have detected an effect at any of these loci.]

2. Is there evidence that any of the SNP loci are associated with variation in more than one of the traits?

3. Given your answer to #2, would you expect to see a change in pupal development time, total development time, or both if natural selection led to a change in pupal mass? Why or why not?

# Short example of an analysis

I use `set.seed(1234)` to ensure that when you run this example you get the same numbers I do. You'll need to change the second to last line to pick the trait and the range of loci that you've been assigned.

```{r message = FALSE}
options(tidyverse.quiet = TRUE)
library(tidyverse)
library(rstan)
library(brms)
library(popkin)

rm(list = ls())

set.seed(1234)

options(mc.cores = parallel::detectCores())

standardize <- function(x) {
  return((x - mean(x, na.rm = TRUE))/sd(x, na.rm = TRUE))
}

analyze_trait <- function(dat, A, trait, loci, n_samples = 2000, n_chains = 4) {
  n_loci <- length(loci)
  p_raw <- matrix(nrow = n_loci, ncol = n_chains*n_samples/2)
  mean_p <- numeric(n_loci)
  mu <- array(dim = c(n_loci, n_samples, nrow(dat)))
  for (i in 1:n_loci) {
    locus <- colnames(dat)[loci[i] + 5]
    cat("Checking locus ", loci[i], " (", locus, ")\n", sep = "")
    fit <- brm(paste(trait, locus, sep = " ~ "),
               data = dat,
               data2 = list(A = A),
               family = gaussian(),
               iter = n_samples,
               refresh = 0)
    x <- as.data.frame(fit)
    brm_locus <- paste("b_", locus, sep ="")
    p_raw[i,] <- x[[brm_locus]]
    mean_p[i] <- mean(x[[brm_locus]])
  }
  loci <- data.frame(locus = colnames(dat)[(1:n_loci) + 5],
                     p_mean = mean_p,
                     index = 1:n_loci)
  loci <- loci[order(abs(loci$p_mean), decreasing = TRUE), ]
  p <- matrix(nrow = n_loci, ncol = n_chains*n_samples/2)
  for (i in 1:n_loci) {
    p[i, ] <- p_raw[loci$index[i],]
  }
  rownames(p) <- loci$locus
  return(list(loci = loci, p = p))
}

summarize_analysis <- function(results) {
  n_report <- nrow(results$loci)
  cat("\n\n",
      "         Mean: (    2.5%,      10%,      50%,      90%,    97.5%)\n")
  dat <- tibble(Locus = character(n_report),
                Mean = numeric(n_report), 
                `2.5%` = numeric(n_report), 
                `10%` = numeric(n_report),
                `50%` = numeric(n_report), 
                `90%` = numeric(n_report), 
                `97.5%` = numeric(n_report))
  for (i in 1:n_report) {
    ci <- quantile(results$p[results$loci$locus[i], ],
                   c(0.025, 0.1, 0.5, 0.9, 0.975))
    output <- sprintf("%6s: %8.5f (%8.5f, %8.5f, %8.5f, %8.5f, %8.5f)\n",
                      results$loci$locus[i],
                      results$loci$p_mean[i],
                      ci[1], ci[2], ci[3], ci[4], ci[5])
    cat(output)
    dat$Locus[i] <- results$loci$locus[i]
    dat$Mean[i] <- results$loci$p_mean[i]
    dat$`2.5%`[i] <- ci[1]
    dat$`10%`[i] <- ci[2]
    dat$`50%`[i] <- ci[3]
    dat$`90%`[i] <- ci[4]
    dat$`97.5%`[i] <- ci[5]
  }
  write_csv(dat, "GWAS-results.csv")
}

## read the data
##
dat <- read_csv("gypsymoth.csv",
                show_col_types = FALSE)
## get the relatedness matrix
##
genos <- dat %>% select(starts_with("X"))
A <- popkin(t(as.matrix(genos)),
            subpops = dat$pops)
## identify the rows with the sample names
##
rownames(A) <- dat$sample

## This example analyzes the first 10 loci.
## Change 1:10 to match the range of loci you have been assigned and "Mass" to the 
## trait you have been assigned 
##
results <- analyze_trait(dat, A, "Mass", 1:10)
summarize_analysis(results)
```

# How I prepared the data for analysis

If you're interested in seeing how I prepared the data from the version on Dryad. Here's what I did. 

First, I downloaded the data from Dryad: [https://doi.org/10.5061/dryad.8ts2867](https://doi.org/10.5061/dryad.8ts2867). Then I renamed the phenotype filename so that it begins with `gm`. The rest is pretty straightforward, except that you haven't seen `popkin()` before, which produces the relatedness matrix. I also fiddle a bit with the results from `popkin()` to get them in the form that's most convenient for analysis in `Stan`.

```{r}
library(tidyverse)
library(popkin)

rm(list = ls())

genos <- read_tsv("gm_genotypes_012_final_5122017.txt",
                  col_names = FALSE,
                  na = "-1",
                  show_col_types = FALSE)
phenos <- read_csv("gm_phenotypes.csv",
                   show_col_types = FALSE)

## keep only loci scored in more than 100 individuals
##
n_scored <- apply(genos, 2, sum, na.rm = TRUE)
genos <- genos[, n_scored > 100]

## keep only individuals scored at more than 4000 loci
##
n_scored <- apply(genos, 1, sum, na.rm = TRUE)
genos <- genos[n_scored > 4000, ]
phenos <- phenos[n_scored > 4000, ]
## and exclude any remaining loci where one or more individuals isn't scored
##
genos <- genos[, !is.na(apply(genos, 2, sum))]

## strip off the individual ids to identify populations
##
pops <- gsub("_.*", "", phenos$sample)

## bind the poopulation, phenotype, and genotype information together and save it in a CSV file
##
dat <- cbind(pops, phenos, genos)
write_csv(dat, 
          file = "gypsymoth.csv")

## estimate the kinship among individuals and save it in a CSV file
##
phi <- popkin(t(as.matrix(genos)), 
              subpops = pops)
L <- t(chol(phi))
write_csv(as.data.frame(L),
          file = "gypsymoth_relatedness.csv")
```
