---
title: "Gila trout workaround"
output: html_notebook
---

## What you need to do

1. Download [gila-trout.zip](http://darwin.eeb.uconn.edu/eeb348-resources/gila-trout.zip)

2. Unizip it in the directory where you want to work. Make sure that it creates a directory called `gila-trout.stru.snmf`. If it does, check to make sure that it has directories K2 through K110. If it does, you should be all set.

3. Download [trout.Rsave](http://darwin.eeb.uconn.edu/eeb348-resources/trout.Rsave) and save it in the directory where you plan to work. `gila-trout.stru.snmf` should be in this directory.

Now you should be able to plot the cross-entropy graph using `plot()` as described in the documentation for `LEA`, and you should be able to use `plot_LEA()` to produce the barplot.

## What I did to produce the files

Here's how I run the data

```{r}
library(LEA)
trout <- snmf("gila-trout.stru.geno", K = 2:10, repetitions = 10, 
              project = "new", 
              entropy = TRUE)
```

I'm going to save the results in an `.Rsave` file. That's what I'll give you, so that you can load it and use the results as if you'd run `snmf()` yourself.

```{r}
save(trout, file = "trout.Rsave")
```

Then I zipped all of the files in the `gila-trout.stru.snmf` directory.