\documentclass[12pt,titlepage,openright]{book}
\usepackage{lecture}
\usepackage{html}
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage[
  bookmarks=true,
  bookmarksopen=true,
  pdftitle={Lecture Notes in Population Genetics},
  pdfauthor={Kent E. Holsinger}]{hyperref}
\usepackage{multicol}
\usepackage{makeidx}

\makeindex

\newcommand{\qtl}{{\tt QTL Cartographer}}
\renewcommand{\myurl}[1]{\htmladdnormallink{#1}{#1}}
\renewcommand\bibname{Literature cited}

% number to subsubsections
\setcounter{secnumdepth}{3}
% and display in Table of Contents
\setcounter{tocdepth}{3}

\begin{document}

\begin{titlepage}

{\Large\bf \noindent Lecture Notes

\noindent in

\noindent Population Genetics

}

\vfill

{\noindent Kent E. Holsinger

\noindent Department of Ecology \& Evolutionary Biology, U-3043

\noindent University of Connecticut

\noindent Storrs, CT  06269-3043}

\vfill

\end{titlepage}

\pagenumbering{roman}

{\small\noindent \copyright{} 2001-2021 Kent E. Holsinger

\ccLicense}

\tableofcontents

\chapter*{Preface}
\addcontentsline{toc}{chapter}{Preface}

\section*{Acknowledgments}

I've used various versions of these notes in my graduate course on
population genetics
\htmladdnormallink{http://darwin.eeb.uconn.edu/eeb348}{http://darwin.eeb.uconn.edu/eeb348}
since 2001. Some of them date back even earlier than that. Several
generations of students, teaching assistants, and colleagues have
found typographical errors, substantive errors, and better ways of
explaining arcane concepts. In addition, the following people have
found various errors and helped me to correct them.

\begin{multicols}{2}
  \noindent Brian Cady \\
  Paul Lewis \\
  Nora Mitchell \\
  Zacharay Muscavitch \\
  Kristen Nolting \\
  Rachel Prunier \\
  Uzay Sezen \\
  Robynn Shannon \\
  Jennifer Steinbachs \\
  Kathryn Theiss \\
  Yufeng Wu \\
\end{multicols}

I am indebted to everyone who has found errors or suggested better
ways of explaining concepts, but don't blame them for any errors that
are left. Those are all mine.

\section*{Note on old chapters}

On the off chance that some readers of these notes may have used parts
of them that I am no longer updating, I've included a section titled
``Old chapters, no longer updated.'' Each of the chapters in this
section has a note at the bottom indicating when it was last
revised. In some cases the material in a chapter has been incorporated
into one of the chapters that is being maintained. In other cases,
I've dropped the material from my course. In either case, the material
doesn't contain any egregious errors (so far as I know), but it will
grow increasingly out of date.

\newpage
\pagenumbering{arabic}
\part{The genetic structure of populations}

\input{genetic-transmission}
\input{hardy-weinberg}
\input{inbreeding}
\input{genetic-structure}
\input{genetic-structure-bayes}
\input{genetic-structure-assignment}

\part{The genetics of natural selection}

\input{selection}
\input{selection-estimate}

\part{Genetic drift}

\input{drift}
\input{mutation-drift}
\input{selection-drift}
\input{coalescent}

\part{Molecular evolution}

\input{molevol-intro}
\input{molevol-patterns}
\input{molevol-selection}

\part{Phylogeography}

\input{molevol-amova}
\input{statistical-phylogeography}
\input{statistical-phylogeography-admixture}
\input{population-genomics}

\part{Quantitative genetics}

\input{quant-intro}
\input{quant-resemblance}
\input{quant-evolution}
\input{association-intro}
\input{genomic-prediction}

\part{Old chapters, no longer updated}

\input{testing-hardy-weinberg}
\input{genetic-structure-hierarchy}
\input{nca}
\input{migrate-ima.tex}
\input{abc}
\input{population-genomics-britain}
\input{two-locus}
\input{selection-components}
\input{selection-multiallele}
\input{association-bamd}
\input{molevol-neutral}
\input{molevol-multigene}
\input{molevol-polymorphism}
\input{molevol-tajima}
\input{mismatch-analysis}
\input{molevol-polymorphism}
\input{molevol-multigene}
\input{winbugs-partitioning}
\input{quant-multiple}
\input{qtl-intro}
\input{qtl-rqtl}
\input{qtl-cartographer}
\input{association-bamd}

\bibliography{popgen}
\bibliographystyle{plain}

\printindex

\end{document}
